---
title: "Analysis of NYC Shootings by Age"
author: "James Roy"
date: "2023-03-03"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Overview

       Known for being the economic capital of the world, New York has a lot of 
    money, but also a lot of crime. Not all crime happens to the same people, 
    at the same time, or for the same reasons. In order to help reduce crime, 
    especially shootings, it is important to look into the data, and figure 
    out where most of these shootings are occurring, why they are happening, and 
    how we can prevent them in the future. 

# Data

           For the scope of this study, we will isolate NYC Precinct Data, and hone in
    on how a person's age affects crime statistics. As of 2006, the NYPD started 
    publicly sharing information on its shootings. The Office of Management 
    Analysis and Planning has oversight on all the data, so it is fairly 
    reliable, but every organization can have its biases.


# Methods and Analysis

          In the next section, we will perform our research and analysis. This will
    provide the reader an in depth guide into the methodology used to derive 
    our results. Ultimately we are going to do the following:
    
    1. Retrieve Data on NYPD Shootings 
    2. Isolate Age Statistics and Shooting Statistics
    3a. Perform Descriptive Analytics, comparing Shooter Age to Victim Age
    3b. Create Graphs to Aide Interpretation of Data
    4. Conclusion, Identification of Bias, and Future Research


*Important Libraries Used: tidyverse, lubridate, stringr*

## Step 1 Identify and Import The Data
      
      In this step we import the main CSV file from the City of New York website. 
      This data includes many variables we will not need, but can save for future study.

```{r libraries}
library(lubridate)
library(stringr)
library(readr)
```

```{r get_nypd_data}

url_in <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")

nypd_data <- url_in
```

## Step 2 Tidy and Transform The Data

      In the following steps we will organize and transform our data. This will
      provide us with clearer understanding of the data set, and help determine how 
      future data will be interacted with. 
      
      For this project, we are going to save four descriptive statistics into our 
      data frame - nypd_data : OCCUR_DATE, BORO, PERP_AGE_GROUP, and VIC_AGE_GROUP. 
      Our main focus will be the latter two variables, which identify the age of the
      perpetrator and the age of the victim.


```{r tidying and transformation}
library(tidyverse)

nypd_data = nypd_data %>% select(OCCUR_DATE, BORO, PERP_AGE_GROUP, 
                                 VIC_AGE_GROUP)

#Filters out unknown information and Outliers

nypd_data_age <- dplyr::filter(nypd_data, PERP_AGE_GROUP %in% 
                                 c("<18","18-24","25-44","45-64","65+"))

# Tidy Data

nypd_data_age

```
## Step 3 Visualize and Analyze

      Now that we have transformed our data, the remaining steps will be easy to follow.
      First we want to observe the perpetrator statistics in a bar graph. Then we do
      the same with the victim statistics. 

```{r Analysis & Visualization}

# Bar graph showing the age of perps in shootings

perp_age_graph <- ggplot(nypd_data_age, aes(x = PERP_AGE_GROUP)) +
  geom_bar() +
  labs(title = "Perpetrator Age",
       x = "Perp Age",
       y = "Shootings")

perp_age_graph

# Bar graph showing the age of victims in shootings

victim_age_graph <- ggplot(nypd_data_age, aes(x = VIC_AGE_GROUP)) +
  geom_bar() +
  labs(title = "Victim Age",
       x = "Victim Age",
       y = "Shootings")

victim_age_graph

```

In this intermediate step, we isolate the vic_age numbers and the perp_age
numbers into a table. This allows us to work with the actual numbers in 
relation to graphical representations.

```{r Analysis Continued }
#Raw Table Showing the numbers of this intersection between perp age and victim age

table(nypd_data_age$PERP_AGE_GROUP,nypd_data_age$VIC_AGE_GROUP)

```

After comparing the Bar Graphs of Victim Age and Perpetrator Age, it is
apparent that the numbers are very similar. We want to figure out why. Next we 
combine the two categories into a count plot. This helps us to visualize where 
the similarities actually are between perpetrator and victim age. 


```{r Analysis & Visualization Continued}

#Countplot showing intersection of perp age and victim age

count_plot <- ggplot(nypd_data_age, aes(y=VIC_AGE_GROUP, x=PERP_AGE_GROUP)) +
  geom_count(alpha=1) +
  labs(title = "Ages",
       x = "Perps",
       y = "Victims",
       size = "Shootings")
count_plot
```


##  Step 4 Analysis and Conclusion

# Graphs

           Analyzing the perpetrator age graph, we see that the majority of shootings are
    committed by people between the ages of 18 and 44. We can also note that people
    under the age of 18 are significantly more likely to shoot a firearm than someone
    over the age of 45.
    
         When we analyze the victim graph, we find the charts eerily similar. People over
    45, are less likely to shoot a gun and they are less likely to be shot by a gun.
    Between the age of 18 and 44, people are more likely to shoot, and be shot in NYC.
    
          Now, we see there's a similarity. We can prove that 18-24 year olds in NYC are 
    more likely to be involved in a shooting, followed by 25-44 year olds, <18 year olds,
    45-64 year olds, and 65+ year olds. We want to map this in a visual way so that
    a correlation can arise. For this we create a type of heat map, called a count plot.
    The count plot shows how 18-24 year olds and 25-44 year olds interact with each other,
    and emphasizes the likelihood of being a victim or a shooter. 
 

# Conclusion

            	The purpose of this study was to determine how age affects the likelihood of a
    gun crime to be committed in NYC. What we found out was rather interesting. First,
    we categorized victims and perps into age groups: <18, 18-24, 25-44, 45-64, and 
    65+. We then transformed the the data using R Tidyverse. Next, we created two 
    bar graphs, a useful data table, and a heat map, representing a correlation
    between shooter age and victim age. With these visualizations and descriptive
    statistics we were able to draw the following conclusions: 1. NYC Persons over
    the age of 45 are unlikely to be involved in a crime. 2. NYC Persons between the
    age of 18 and 44, are most likely to to be involved in a crime, whether the 
    perpetrator or the victim. 3.It is likely that crimes are typically committed 
    within specific age groups. We know this because of the large number of criminals
    and victims in the 18-44 range.



# Bias

          In any data analysis, it is impossible to rid yourself of all biases. We have done
    our best to remain as objective as possible, only presenting the data and findings
    that we have through this R document. The steps and methodology are all available.
    Despite this, it is possible the NYC government wants to undercut their crime 
    statistics. It is also possible that certain age groups do not report as many shootings.
    Furthermore, we removed some data that is unknown, since there are 
    ongoing cases and investigations. There are many biases to be wary of, but all in all
    we have done a good job at deterring and removing any forms of bias. 

# Future Research

          In the future we would like to study some of the other statistics that affect
    shooting crime, such as income, precinct, education, nationality, sex, housing,
    and so on. 
    
          The current step is creating a model to determine exactly how the age of the 
    victim is related to the age of the perp. As of now we know who is most likely
    to commit a shooting, and who is most likely to be shot. But we do not know 
    precisely why this is true, or who exactly they would be shot by. We can speculate,
    however, that shootings typically happen within age groups.



